---
title: "This Is Not the ex-Gaussian Model You Are Looking For: On the Default Parameterization of Bayesian ex-Gaussian Models in brms"
shorttitle: "Ex-Gaussian Parameterization in brms"
author:
  - name: Jason Geller
    corresponding: true
    orcid: 0000-0002-7459-4505
    email: drjasongeller@gmail.com
    affiliations:
      - name: Boston College
        department: Department of Psychology and Neuroscience
        address: McGuinn 300
        city: Chestnut Hill
        region: MA
        country: USA
        postal-code: 02467
    roles:
      - Conceptualization
      - Data curation
      - Formal analysis
      - Project administration
      - Resources
      - Visualization
      - Writing - original draft
  - name: Bernhard Angele
    orcid: 0000-0001-8989-8555
    affiliations:
      - name: Nebrija University
    roles:
     - Formal analysis
     - Validation
     - Writing - original draft
     - Writing - review & editing
  - name: Dominique Makowski
    affiliations:
      - name: University of Sussex
        department: School of Psychology
        country: UK
    roles:
      - Formal analysis
      - Validation
      - Software
      - Writing - review & editing
abstract:
  The ex-Gaussian distribution is a popular mathematical model for analyzing response time
  (RT) data because it separates the central portion of the distribution,
  captured by the Gaussian parameters $\mu$ and $\sigma$, from the positively
  skewed tail, captured by the exponential parameter $\tau$. Researchers often use these
  parameters to make inferences about underlying cognitive processes. As Bayesian methods
  have become increasingly accessible, complex ex-Gaussian models can now be fit
  with relative ease, and in R, the {brms} package makes this especially
  straightforward. However, it is important to recognize that the default
  parameterization offered by {brms} does not map onto the classical
  parameterization familiar to many researchers in experimental psychology in the
  default model, $\mu$ indexes the mean of the full ex-Gaussian distribution rather
  than the location of the Gaussian component alone. This distinction matters
  because changes in the Gaussian location and changes in the exponential tail
  can offset one another at the level of the overall mean, so some published
  studies may have drawn incorrect inferences from effects estimated on the
  default $\mu$ parameter. Using simulated data and a published finding we highlight 
  how using the default parameterization can lead to misleading inferences. We then demonstrate 
  how to fit the classical parameterization directly using the {cogmod} package, with the aim of helping researchers 
  and reviewers align ex-Gaussian model parameters with their theoretical questions.
keywords:
  - ex-Gaussian
  - reaction time
  - Bayesian estimation
  - brms
  - parameterization
citeproc: true
floatsintext: true
wordcount: true
author-note:
  disclosures:
    study-registration: "The analyses herein were not preregistered."
    data-sharing: Data, code, and materials for this manuscript are available at https://github.com/jgeller112/exGauss_commentary. The entire manuscript, including all analyses and figures, can be reproduced by downloading the repository and following the instructions in its README.
    conflict-of-interest: The authors have no conflicts of interest to disclose.
    financial-support: ~
    authorship-agreements: ~
numbered-lines: true
bibliography: references.bib
date: "`r Sys.Date()`"
knitr:
  opts_chunk:
    echo: true
    results: hide
    message: false
    warning: false
    freeze: auto
    dev: "ragg_png"
    fig-align: center
    tbl-align: center
    code-line-numbers: true
    keep-with-next: true
    code-overflow: wrap
    fig-width: 6
    fig-asp: 0.8
    fig-pos: "h"
    tbl-pos: "h"
    fig-dpi: 300
format:
  apaquarto-typst:
    csl: apa.csl
  apaquarto-docx: default
  apaquarto-html:
    toc: true
    toc-depth: 2
    toc-title: Manuscript Outline
    toc-location: right
    echo: true
    code-tools: true
    code-fold: true
    code-summary: "Show R code"
---

```{r}
#| label: load-packages
#| echo: false
#| message: false
#| warning: false
library(tidyverse)
library(fs)
library(brms)
library(ragg)
library(tinytable)
library(emmeans)
library(bayestestR)
library(parameters)
library(modelbased)
library(ggplot2)
library(ggdist)
library(qs2)
library(patchwork)
library(cogmod)
```

Over the past two decades, psychology has seen a substantial rise in the application of Bayesian data analysis methods as a supplement to widely used frequentist methods [@pfadt2025methodological]. This shift has been driven by the increasing availability of computational tools that have made fitting Bayesian models both practical and efficient. Programming languages like R [@R], Python [@python], and Julia [@julia] provide libraries and packages that allow researchers to fit Bayesian models with little technical overhead. In addition, growing dissatisfaction with the null ritual and its associated pitfalls (e.g., p-value misinterpretation and dichotomous thinking) [@gigerenzer2018statistical], together with growing distrust of published findings [@vazire2018], has pushed many researchers toward statistical alternatives.

One domain that has seen increased adoption of Bayesian methods is experimental psychology, where one of the most common dependent variables is response times (RT), which is a measure of how long it takes a person to initiate some type of response in the presense of a stimulus (i.e., motor, written, or spoken). RT data offers a window into cognitive processes undelying behavior [@baayen2010analyzing]. Traditionally, RT data have been analyzed using summary statistics such as the mean or median. However, RT distributions are typically positively skewed, bounded at zero, and characterized by substantial variability both within and between individuals [@wilcox; @whelan2008]. Moreover, experimental manipulations may affect not only the central tendency of the distribution but also its variability or upper tail. Consequently, analyses based solely on the mean or median can obscure theoretically meaningful distributional effects, motivating calls to move beyond mean-based analyses of RT data [@balota2011].

One way to move beyond the mean is to model the full shape of the RT distribution. Mathematical models provide a flexible way to do so, allowing researchers to draw inferences about the psychological processes—such as memory, attention, and language—that give rise to observed response times. One model that has become especially popular for this purpose is the ex-Gaussian model [@ratcliff1976]. The ex-Gaussian distribution is formed by the convolution of a Gaussian distribution and an exponential distribution and is characterized by three parameters: $\mu$, $\sigma$, and $\tau$. The parameter $\mu$ represents the mean of the Gaussian component, whereas $\sigma$ represents its standard deviation. The $\tau$ parameter represents both the mean and standard deviation of the exponential component and primarily governs the length of the distribution’s positively skewed upper tail. The overall mean of the ex-Gaussian distribution is $\mu$ + $\tau$ , and the standard deviation (SD) is $\sqrt{\sigma^2+\tau^2}$ . Although these parameters have been used to infer cognitive processes such as automaticity [@dewitt2018; @balota2008] and cognitive control/inhibition [@kane2003], these interpretations are not inherent to the parameters and should be made cautiously [@matzke2009]. Nevertheless, the ex-Gaussian model can provide a more complete description of RT data than the arithmetic mean alone, which can be disproportionately influenced by the positively skewed tail of the distribution.

The ex-Gaussian model can be fit within a Bayesian framework relatively easily using the {brms} package [@brms2017; @brms2018; @brms2021] in R which provides a high-level interface to the probablitsic programming language Stan [@stan]. The {brms} package supports a wide range of response distributions, including the ex-Gaussian family, and allows researchers to incorporate complex model structures at both the participant and item levels as well as across different parameters. This accessibility has contributed to the growing use of Bayesian ex-Gaussian models in reaction-time research [see @angele2023; @angele2024reading; @geller2025; @kinoshita2023; @serranocarot2026]. However, an important complication arises from the default parameterization used by the R package {brms}. Although the model is straightforward to estimate, its default specification does not necessarily provide direct access to the parameter-specific comparisons researchers often wish to make.

The central problem is that the parameter labeled $\mu$ in the default {brms} ex-Gaussian model does not correspond to the mean of the Gaussian component in the conventional parameterization. Instead, the quantity modeled as $\mu$ in {brms} represents the mean of the full ex-Gaussian distribution. Under the conventional parameterization, this quantity is equal to $\mu_{\mathrm{Gaussian}} + \tau$. Consequently, regression effects estimated for the default $\mu$ parameter reflect changes in a composite quantity that combines the Gaussian location and exponential tail components. Such effects therefore cannot be interpreted directly as changes in the Gaussian component or in typical response speed. To recover the conventional Gaussian mean, it is necessary to derive it from the model parameters as $\mu_{\mathrm{Gaussian}} =
\mu_{\mathrm{brms}} - \tau$ or to specify an alternative parameterization that models the conventional ex-Gaussian parameters directly.

In this commentary, we demonstrate what the default parameterization of the ex-Gaussian model in {brms} actually provides and then show how to specify the model so that its parameters correspond to the conventional ex-Gaussian parameters used in reaction-time research. The goal is to improve the accuracy and interpretability of inferences drawn from these models. This issue is particularly important because the ex-Gaussian distribution is widely used to distinguish changes in the central portion of the RT distribution from changes in its positively skewed tail.

## Simulation Example

### Data Generation

To demonstrate the issue with fitting the default ex-Gaussian model in {brms}, we simulate data using a comparison adapted from one of our own studies [@geller2025] in R [version 4.6.1, @R][^1]. In @geller2025, we examined how perceptual disfluency—manipulated by varying the level of blur applied to letter strings influenced reaction times in a lexical decision task (LDT; i.e., is this letter string a word?), and whether these effects were related to how well someone remembers the words in a recognition memory task. In that study, low blur words affected early stages of encoding, reflected in the Gaussian location parameter $\mu$, but not the later, tail component captured by $\tau$. For the present illustration, however, we construct a sharper scenario in which $\mu$ and $\tau$ differ in opposite directions so that the two conditions share the same expected reaction time. Using the classical ex-Gaussian parameterization implemented in the {cogmod} package [@makowski2026cogmod], we simulate ex-Gaussian data for clearly presented and low-blur words in which $\mu$ is set to 500 ms and 550 ms, $\tau$ to 100 ms and 50 ms, and $\sigma$ to 60 ms in both conditions (see @tbl-exgauss).

[^1]: We used the following R packages: bayestestR v. 0.18.1 [@bayestestR], brms v. 2.23.0 [@brms2017; @brms2018; @brms2021], cogmod v. 0.0.21 [@makowski2026cogmod], emmeans v. 2.0.4 [@emmeans], fs v. 2.1.0 [@fs], ggdist v. 3.3.3 [@ggdist2024; @ggdist2025], here v. 1.0.2 [@here], modelbased v. 0.16.0 [@makowski2025modelbased], parameters v. 0.29.2 [@parameters], patchwork v. 1.3.2 [@patchwork], qs2 v. 0.2.2 [@qs2], ragg v. 1.5.2 [@ragg], rmarkdown v. 2.31 [@rmarkdown2018; @rmarkdown2020; @rmarkdown2026], tidyverse v. 2.0.0 [@tidyverse], tinytable v. 0.17.0 [@tinytable].

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false

# Check if split .qs2 files need to be combined (e.g. after a fresh git clone)
split_qs2_script <- if (file.exists("split_qs2.R")) "split_qs2.R" else if (file.exists("../split_qs2.R")) "../split_qs2.R" else NULL
if (!is.null(split_qs2_script)) {
  source(split_qs2_script)
  # _quarto.yml only exists at the project root, so this is unambiguous
  # (unlike checking for Angele_et_al_2022/, which can also exist under ms/
  # locally e.g. from old interactive-session artifacts).
  root_dir <- if (file.exists("_quarto.yml")) "." else ".."
  check_and_combine_qs2(root_dir)
}

# Where brm(..., file = ...) below caches fitted models across renders.
# root_dir already points from the current working directory (either the
# project root or ms/, depending on execute-dir) back to the project root,
# so appending "ms/fits" resolves correctly either way.
fits_dir <- file.path(root_dir, "ms", "fits")
if (!dir.exists(fits_dir)) dir.create(fits_dir, recursive = TRUE)

# Shared formatting helpers, reused when building the condensed (long-format)
# tables: one row per condition/effect x parameter, with Greek symbols (μ, τ,
# β) as values in a "Parameter" column rather than as column headers.
fmt_num <- function(x, digits = 1) {
  formatC(round(x, digits), format = "f", digits = digits)
}
fmt_ci <- function(low, high, digits = 1) {
  paste0("[", fmt_num(low, digits), ", ", fmt_num(high, digits), "]")
}
fmt_pd <- function(x, digits = 3) {
  formatC(round(x, digits), format = "f", digits = digits)
}
```

```{r}
#| label: load-brms-exgaussian
#| lst-label: lst-install-brms-exgaussian
#| lst-cap: "Installing and loading the {cogmod} package from GitHub."
#| eval: false

# install from Github
#remotes::install_github("DominiqueMakowski/cogmod")

library(cogmod)
```

```{r}
#| label: tbl-exgauss
#| tbl-cap: "Data-generating ex-Gaussian parameters for the Clear and Low-Blur conditions, with opposing changes in $\\mu$ and $\\tau$ that hold the expected response time ($\\mu+\\tau$) constant."
#| echo: false
#| results: asis

# Data-generating ex-Gaussian parameters for each Blur condition:
#   mu    = location of the Gaussian component
#   sigma = SD of the Gaussian component
#   tau   = mean of the exponential component (the skewed tail)
sim_params <- data.frame(
  Blur = c("Clear", "Low-Blur"),
  mu = c(500, 550),
  sigma = c(60, 60),
  tau = c(100, 50)
)

# Expected RT of an ex-Gaussian is mu + tau; shown as a reference column.
sim_params$full_mean <- sim_params$mu + sim_params$tau

# Relabel columns with Greek symbols for display.
param_table <- sim_params
colnames(param_table) <- c("Blur", "μ", "σ", "τ", "Expected RT")

tt(param_table)
```

To simulate the data, We first draw 100 reaction times per condition (200 in total) from the ex-Gaussian distributions defined by these parameters (@lst-simulate-data).

```{r}
#| label: simulate-data
#| lst-label: lst-simulate-data
#| lst-cap: "Simulating ex-Gaussian reaction times for the Clear and Low-Blur conditions."
# Fix the random seed so the simulated data are identical on every render.
set.seed(1234)

# Number of simulated trials per condition (200 observations in total).
n_per_group <- 100

# One row per trial, labelled by Blur condition (Clear is the reference level).
d <- data.frame(
  Blur = factor(
    rep(c("Clear", "Low-Blur"), each = n_per_group),
    levels = c("Clear", "Low-Blur")
  )
)

# Row index (1 = Clear, 2 = Low-Blur) used to look up each trial's parameters.
parameter_index <- as.integer(d$Blur)

# Draw reaction times from the classical ex-Gaussian using each trial's
# condition values, via {cogmod}'s rrt_exgaussian() generator.
d$rt <- cogmod::rrt_exgaussian(
  n = nrow(d),
  mu = sim_params$mu[parameter_index],
  sigma = sim_params$sigma[parameter_index],
  tau = sim_params$tau[parameter_index]
)
```

@fig-datagen shows the simulated RT distribution for each condition and how the three ex-Gaussian parameters shape it: $\mu$ locates the Gaussian component, $\sigma$ sets its spread, and $\tau$ stretches the right tail. Because the increase in $\mu$ from Clear to Low-Blur is offset by an equal decrease in $\tau$, the overall mean $E(RT) = \mu + \tau$ is identical (600 ms) in both conditions.

```{r}
#| label: fig-datagen
#| fig-cap: "Simulated reaction-time distributions for each condition, annotated with the ex-Gaussian parameters. The dashed line marks the Gaussian location $\\mu$; the double arrow spans $\\pm\\sigma$ around it (the Gaussian spread); the rightward arrow over the tail marks the exponential component $\\tau$, which stretches the distribution to the right; and the solid line marks the overall mean $E(RT) = \\mu + \\tau$, which is 600 ms in both conditions. The bottom span highlights that $E(RT)$ is the mean of the *entire* distribution — the quantity the default `brms` `mu` estimates."
#| echo: false
#| fig-width: 6.5
#| fig-height: 4.7

# E(RT) = mu + tau, used to place the annotations.
annot <- sim_params
# Horizontal extent used for the "entire distribution" span.
annot$lo <- annot$mu - 2.5 * annot$sigma
annot$hi <- annot$full_mean + 4 * annot$tau

ggplot(d, aes(x = rt)) +
  # RT distribution for each condition (kernel density via ggdist).
  ggdist::stat_slab(
    aes(fill = Blur),
    normalize = "panels",
    color = NA,
    alpha = 0.85
  ) +
  # mu: Gaussian location (dashed) and E(RT) = mu + tau (solid).
  geom_vline(
    data = annot,
    aes(xintercept = mu),
    linetype = "dashed",
    linewidth = 0.5,
    colour = "grey25"
  ) +
  # Bounded segment (not geom_vline) so the solid E(RT) line stops before the
  # "entire distribution" annotation below the panel instead of running through it.
  geom_segment(
    data = annot,
    aes(x = full_mean, xend = full_mean, y = -0.12, yend = 1.3),
    linewidth = 0.5,
    colour = "black"
  ) +
  # sigma: double arrow spanning +/- sigma around mu (white, on the coloured fill).
  # Placed low (y = 0.22) so both arrowheads stay inside the fill for both
  # conditions, including the narrower Low-Blur distribution.
  geom_segment(
    data = annot,
    aes(x = mu - sigma, xend = mu + sigma, y = 0.22, yend = 0.22),
    arrow = grid::arrow(length = grid::unit(1.6, "mm"), ends = "both"),
    linewidth = 0.5,
    colour = "white"
  ) +
  geom_text(
    data = annot,
    aes(x = mu, y = 0.33, label = "sigma"),
    parse = TRUE,
    size = 4.4,
    fontface = "bold",
    colour = "white"
  ) +
  # tau: rightward arrow over the tail (the exponential component stretches it right).
  geom_segment(
    data = annot,
    aes(x = full_mean, xend = full_mean + 2 * tau, y = 1.06, yend = 1.06),
    arrow = grid::arrow(length = grid::unit(1.9, "mm"), ends = "last"),
    linewidth = 0.5,
    colour = "grey20"
  ) +
  geom_text(
    data = annot,
    aes(x = full_mean + tau, y = 1.16, label = "tau~(tail)"),
    parse = TRUE,
    size = 4.4,
    fontface = "bold"
  ) +
  # axis-side labels for mu and E(RT) (nudged apart so they don't overlap).
  geom_text(
    data = annot,
    aes(x = mu, y = -0.09, label = "mu"),
    parse = TRUE,
    size = 4.6,
    fontface = "bold",
    colour = "grey20",
    hjust = 1,
    nudge_x = -8
  ) +
  geom_text(
    data = annot,
    aes(x = full_mean, y = -0.09, label = "E(RT)"),
    size = 4,
    fontface = "bold",
    hjust = 0,
    nudge_x = 8
  ) +
  # entire distribution: full-width span (its mean is E(RT); see caption).
  geom_segment(
    data = annot,
    aes(x = lo, xend = hi, y = -0.19, yend = -0.19),
    arrow = grid::arrow(length = grid::unit(1.4, "mm"), ends = "both"),
    linewidth = 0.4,
    colour = "grey45"
  ) +
  geom_text(
    data = annot,
    aes(x = full_mean, y = -0.55, label = "entire distribution"),
    size = 3.3,
    colour = "grey30"
  ) +
  facet_wrap(~Blur, ncol = 1) +
  scale_fill_manual(
    values = c("Clear" = "darkgreen", "Low-Blur" = "darkred"),
    guide = "none"
  ) +
  coord_cartesian(ylim = c(-0.38, 1.28), xlim = c(250, 1150), clip = "off") +
  labs(x = "Reaction time (ms)", y = NULL) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.y = element_blank(),
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold"),
    plot.margin = margin(t = 5.5, r = 5.5, b = 22, l = 5.5)
  )
```

### Default {brms} Parameterization

Using this data we fit the default `brms` ex-Gaussian model. In this parameterization, the parameter labeled `mu` represents the expected value of the full ex-Gaussian distribution rather than the location of its Gaussian component.

The exponential component is labeled `beta`. On the response scale, `beta` corresponds to the classical ex-Gaussian parameter $\tau$:

$$\beta_{\texttt{brms}} = \tau_{\text{classical}}.$$

Thus, `beta` and $\tau$ are not substantively different quantities. Both represent the mean of the exponential component. The difference is primarily one of notation. The important difference between the default {brms} model and the classical parameterization concerns `mu`:

$$\mu_{\texttt{brms}} = \mu_{\text{classical}} + \tau.$$

In other words, the default {brms} parameterization uses `mu` for the expected value of the complete ex-Gaussian distribution, whereas the classical parameterization uses $\mu$ for the location of the Gaussian component.

Both the expected response and the exponential component are allowed to vary across conditions.

#### Prior Specification

The priors were selected to reflect the known scale of the simulated data while remaining sufficiently broad to allow the model to recover deviations from the generating values. @tbl-priors shows these priors alongside those used for the classical parameterization introduced later, so the two specifications can be compared directly.

```{r}
#| label: priors-default
#| echo: false
priors_default <- c(
  # mu intercept: centered at 600 ms, the expected RT (E(RT)) of the reference
  # condition in this default parameterization.
  prior(
    normal(600, 100),
    class = "Intercept"
  ),
  # Effect of Blur on mu: centered at 0 (no assumed direction).
  prior(
    normal(0, 50),
    class = "b"
  ),
  # beta (exponential mean) intercept, on the log scale: exp(log(100)) = 100 ms.
  prior(
    normal(log(100), 0.5),
    class = "Intercept",
    dpar = "beta"
  ),
  # Effect of Blur on beta (log scale): centered at 0 (a ratio of 1).
  prior(
    normal(0, 0.5),
    class = "b",
    dpar = "beta"
  )
)
```

```{r}
#| label: tbl-priors
#| tbl-cap: "Prior specification for the default and classical ex-Gaussian models."
#| echo: false
#| results: asis
priors_table <- data.frame(
  Parameter = c(
    "Intercept (μ)",
    "Blur (μ)",
    "Intercept (β / τ)",
    "Blur (β / τ)",
    "Intercept (σ)"
  ),
  Default = c(
    "Normal(600, 100)",
    "Normal(0, 50)",
    "Normal(log(100), 0.5)",
    "Normal(0, 0.5)",
    "brms default"
  ),
  Classical = c(
    "Normal(500, 100)",
    "Normal(0, 50)",
    "Normal(log(100), 0.5)",
    "Normal(0, 0.5)",
    "Normal(log(60), 0.5)"
  ),
  Notes = c(
    "Same form; centered on each parameterization's reference-condition value (E(RT) = 600 ms vs. Gaussian location = 500 ms); both mu and E(RT) use an identity link, so both priors are on the response scale",
    "Identical across parameterizations; no assumed direction for the effect of Blur",
    "Identical form in both models: brms's native `exgaussian()` and {cogmod}'s `rt_exgaussian()` (here fit with `link_tau = \"log\"`) both use a log link for the exponential component, so the prior is on the log scale, centered at log(100 ms)",
    "Identical across parameterizations; no assumed direction for the effect of Blur, expressed on the log scale",
    "Only the classical model estimates sigma explicitly; fit with `link_sigma = \"log\"`, so the prior is on the log scale, centered at log(60 ms)"
  )
)
colnames(priors_table) <- c(
  "Parameter",
  "Default Prior",
  "Classical Prior",
  "Notes"
)

tt(priors_table)
```

The model can then be fit as follows (@lst-fit-default):

```{r}
#| label: fit-default
#| lst-label: lst-fit-default
#| lst-cap: "Fitting the default ex-Gaussian model in brms."
fit_brms <- brm(
  bf(
    rt ~ Blur, # mu submodel: here mu is E(RT), not the Gaussian location
    sigma ~ 1, # constant Gaussian SD
    beta ~ Blur # exponential mean (the classical tau), allowed to vary
  ),
  data = d,
  prior = priors_default,
  family = exgaussian(), # brms' built-in ex-Gaussian family
  chains = 2,
  iter = 2000,
  warmup = 500,
  backend = "cmdstanr",
  cores = 4,
  seed = 1234, # reproducible draws
  file = file.path(fits_dir, "fit-default"), # cache: skip refitting if unchanged
  file_refit = "on_change" # but do refit if the model spec above changes
)
```

#### Estimating Ex-Gaussian Parameters with `{easystats}`

The `{modelbased}` package [@makowski2025modelbased], which is part of the `{easystats}` ecosystem [@ludecke2019insight], can easily be used to obtain condition-specific posterior estimates from a fitted model. The `estimate_means()` function returns estimated parameters for each condition, back-transformed, so estimates are presented on time scale, whereas `estimate_contrasts()` computes differences between conditions.

Throughout this commentary, we treat an effect as reflecting a credible difference when its 95% credible interval excludes zero and its probability of direction (pd) exceeds .975; when either condition fails to hold, we describe the effect as non-credible or uncertain.

As shown in @tbl-exgauss, the present data were generated so that the two conditions have the same expected reaction time but differ in the Gaussian and exponential components of the ex-Gaussian distribution.

Under the classical ex-Gaussian parameterization, the expected response is

$$E(RT) = \mu + \tau.$$

Thus, the 50-ms increase in $\mu$ for the Low-Blur condition is offset by a 50-ms decrease in $\tau$. The expected RT is therefore 600 ms in both conditions.

#### Estimating the Expected Response ($\mu$)

Condition-specific estimates of the expected response can be obtained with `estimate_means()` and setting `predict` argument to "mu".

```{r}
#| label: estimate-mu-default
#| echo: false
# Default model: per-condition mu, which equals E(RT) in this parameterization.
overall_means <- estimate_means(
  fit_brms,
  by = "Blur",
  predict = "mu",
  centrality = "mean"
)
```

Because the primary `mu` parameter in the default ex-Gaussian model represents $E(RT)$, these estimates should be close to 600 ms in both conditions.

The corresponding contrast can be obtained using `estimate_contrasts()`.

```{r}
#| label: contrast-mu-default
#| echo: false
# Default model: Blur contrast in E(RT) (≈ 0 by design).
overall_difference <- estimate_contrasts(
  fit_brms,
  contrast = "Blur",
  centrality = "mean"
)
```

The posterior estimate of this contrast should be close to zero because the two conditions were generated with the same expected RT:

$$(550 + 50) - (500 + 100) = 600 - 600 = 0.$$

#### Estimating the Exponential Component

In the default {`brms}` parameterization, `beta` represents the mean of the exponential component. It therefore corresponds directly to $\tau$ in the classical parameterization:

$$\beta = \tau.$$

Condition-specific estimates can be obtained by requesting predictions for `beta`.

```{r}
#| label: estimate-beta-default
#| echo: false
# Default model: per-condition beta (the exponential mean = classical tau).
tau_estimates_default <- estimate_means(
  fit_brms,
  by = "Blur",
  predict = "beta",
  centrality = "mean"
)
```

Although the resulting object is called `tau_estimates_default` for interpretive clarity, the parameter is named `beta` inside the default `brm` model. @tbl-default-means shows both the expected response and the exponential component together, by condition.

```{r}
#| label: build-tbl-default-means
#| echo: false
default_means_mu <- as.data.frame(overall_means)
default_means_beta <- as.data.frame(tau_estimates_default)

default_means_tbl <- rbind(
  data.frame(
    Blur = default_means_mu$Blur,
    Parameter = "E(RT)",
    Mean = fmt_num(default_means_mu$Mu),
    `95% CrI` = fmt_ci(default_means_mu$CI_low, default_means_mu$CI_high),
    pd = fmt_pd(default_means_mu$pd),
    check.names = FALSE
  ),
  data.frame(
    Blur = default_means_beta$Blur,
    Parameter = "β",
    Mean = fmt_num(default_means_beta$Beta),
    `95% CrI` = fmt_ci(default_means_beta$CI_low, default_means_beta$CI_high),
    pd = fmt_pd(default_means_beta$pd),
    check.names = FALSE
  )
)
default_means_tbl <- default_means_tbl[order(default_means_tbl$Blur), ]
rownames(default_means_tbl) <- NULL
default_means_tbl$Blur <- as.character(default_means_tbl$Blur)
default_means_tbl$Blur[duplicated(default_means_tbl$Blur)] <- ""
```

```{r}
#| label: tbl-default-means
#| tbl-cap: "Estimated expected response, $E(RT)$, and exponential component (`beta`), by condition from the default brms model."
#| echo: false
#| results: asis
tt(default_means_tbl)
```

These estimates should be close to 600 ms for $E(RT)$ in both conditions, and approximately 100 ms for Clear and 50 ms for Low-Blur for `beta`.

The corresponding contrasts can be obtained using `estimate_contrasts()`.

```{r}
#| label: contrast-beta-default
#| echo: false
# Default model: Blur contrast in beta (the exponential component).
tau_difference_default <- estimate_contrasts(
  fit_brms,
  contrast = "Blur",
  predict = "beta",
  centrality = "mean"
)
```

```{r}
#| label: build-tbl-default-contrast
#| echo: false
default_contrast_mu <- as.data.frame(overall_difference)
default_contrast_beta <- as.data.frame(tau_difference_default)

default_contrast_tbl <- rbind(
  data.frame(
    Contrast = paste(
      default_contrast_mu$Level1,
      "-",
      default_contrast_mu$Level2
    ),
    Parameter = "E(RT)",
    Estimate = fmt_num(default_contrast_mu$Mean),
    `95% CrI` = fmt_ci(default_contrast_mu$CI_low, default_contrast_mu$CI_high),
    pd = fmt_pd(default_contrast_mu$pd),
    check.names = FALSE
  ),
  data.frame(
    Contrast = paste(
      default_contrast_beta$Level1,
      "-",
      default_contrast_beta$Level2
    ),
    Parameter = "β",
    Estimate = fmt_num(default_contrast_beta$Mean),
    `95% CrI` = fmt_ci(
      default_contrast_beta$CI_low,
      default_contrast_beta$CI_high
    ),
    pd = fmt_pd(default_contrast_beta$pd),
    check.names = FALSE
  )
)
```

```{r}
#| label: tbl-default-contrast
#| tbl-cap: "Blur contrast in the expected response, $E(RT)$, and the exponential component (`beta`), from the default brms model."
#| echo: false
#| results: asis
tt(default_contrast_tbl)
```

As shown in @tbl-default-contrast, the default model correctly recovers a near-zero contrast in $E(RT)$, since the two conditions were generated with the same expected RT. However, its primary `mu` coefficient does not directly reveal the 50-ms difference in the Gaussian-location parameter — that difference only shows up in the `beta` contrast, which is approximately 50 ms.

### Classical Ex-Gaussian Parameterization

As we have shown above, the default paramterization gives you the exected RT of the entire distribution. We can instead fit the model using the `rt_exgaussian()` family from the {cogmod} package, which implements the classical ex-Gaussian parameterization. In this model, $\mu$ represents the location of the Gaussian component and $\tau$ represents the mean of the exponential component.

By default, {cogmod} links all three parameters with a softplus function, rather than the identity or log links {brms} typically applies. An identity link would let the linear predictor cross zero, which is invalid for all three parameters, since $\mu$ must stay positive to represent a genuine RT-scale location, and $\sigma$ and $\tau$ are positive by definition. Softplus keeps the positivity constraint but behaves almost linearly away from zero, which is convenient in principle — but only in the positive direction, and only once the linear predictor is comfortably greater than zero: $\text{softplus}(x) = \log(1+e^x) \to x$ as $x \to \infty$, but $\text{softplus}(x) \to 0$ as $x \to -\infty$, so the link is sharply nonlinear for negative or near-zero linear predictors (e.g., $\text{softplus}(-2) \approx 0.13$, nothing like $-2$). Whether a response-scale prior actually lands in that near-linear region depends on the measurement scale of the outcome, not just the link function itself — the same nominal prior can behave very differently depending on whether an RT is coded in milliseconds or seconds.

Because of this scale-sensitivity, and because we would rather interpret every coefficient directly in `estimate_means()`/`estimate_contrasts()` output than reason about a link function at all, we do not use {cogmod}'s default softplus link in this paper. `rt_exgaussian()` accepts `link_mu`, `link_sigma`, and `link_tau` arguments so each parameter's link can be set independently; we use an identity link for $\mu$ and a log link for $\sigma$ and $\tau$ — the same links {brms}'s own built-in `exgaussian()` family uses for its location and scale parameters, and a familiar, scale-robust choice for enforcing positivity (unlike softplus, log's behavior does not depend on whether the parameter values happen to be small or large in absolute terms). The model is specified with `family = rt_exgaussian(link = "identity", link_sigma = "log", link_tau = "log")`.

At the sampling level, the log-likelihood is written in Stan and calls the built-in `exp_mod_normal_lpdf()` function[^2], which implements the exponentially modified Gaussian density. That built-in function is parameterized by the rate of the exponential component rather than its mean, so $\tau$ is converted internally via $1/\tau$ before being passed in. The custom family simply relabels and re-links Stan's existing exponentially modified Gaussian implementation.

[^2]: <https://github.com/DominiqueMakowski/cogmod/blob/4b10ea5e00e02984d39c0749adf8dc0b9c4ab15b/R/rt_exgaussian_brms.R#L30-L37>

#### Prior Specification

The priors for the classical model follow the same reasoning as those for the default model, shown together in @tbl-priors:

```{r}
#| label: priors-classical
#| echo: false
priors_classical <- c(
  # mu intercept: centered at 500 ms, now the Gaussian location of the
  # reference condition (not E(RT)). mu uses an identity link, so this
  # prior is specified directly on the response (ms) scale.
  prior(
    normal(500, 100),
    class = "Intercept"
  ),
  # Effect of Blur on mu (Gaussian location): centered at 0.
  prior(
    normal(0, 50),
    class = "b"
  ),
  # tau (exponential mean) intercept, on the log scale: exp(log(100)) = 100 ms.
  prior(
    normal(log(100), 0.5),
    class = "Intercept",
    dpar = "tau"
  ),
  # Effect of Blur on tau (log scale): centered at 0.
  prior(
    normal(0, 0.5),
    class = "b",
    dpar = "tau"
  ),
  # sigma (Gaussian SD) intercept, on the log scale: exp(log(60)) = 60 ms.
  prior(
    normal(log(60), 0.5),
    class = "Intercept",
    dpar = "sigma"
  )
)
```

The model is then fit using (@lst-fit-classical):

```{r}
#| label: fit-brms-classical
#| lst-label: lst-fit-classical
#| lst-cap: "Fitting the classical ex-Gaussian model using {cogmod}."
fit_brms_new <- brm(
  bf(
    rt ~ Blur, # mu submodel: here mu IS the Gaussian location
    tau ~ Blur, # exponential mean (the tail), allowed to vary
    sigma ~ 1 # constant Gaussian SD
  ),
  data = d,
  prior = priors_classical,
  family = rt_exgaussian(link = "identity", link_sigma = "log", link_tau = "log"), # {cogmod} family with identity/log links
  stanvars = rt_exgaussian_stanvars(), # Stan code injected into the model
  chains = 2,
  iter = 2000,
  warmup = 500,
  backend = "cmdstanr",
  cores = 4,
  seed = 1234,
  file = file.path(fits_dir, "fit-classical"), # cache: skip refitting if unchanged
  file_refit = "on_change" # but do refit if the model spec above changes
)
```

#### Estimating the Gaussian-Location Parameter

For the classical model, estimates from the primary regression formula correspond to the Gaussian-location parameter $\mu$.

```{r}
#| label: estimate-mu-classical
#| echo: false
# Classical model: per-condition mu, now the Gaussian-location parameter.
mu_estimates <- estimate_means(
  fit_brms_new,
  by = "Blur",
  predict = "mu",
  centrality = "mean"
)
```

The estimates are approximately 500 ms for Clear and 550 ms for Low-Blur.

The contrast between the two can be obtained using `estimate_contrasts()`.

```{r}
#| label: contrast-mu-classical
#| echo: false
# Classical model: Blur contrast in the Gaussian-location parameter mu.
mu_difference <- estimate_contrasts(
  fit_brms_new,
  contrast = "Blur",
  predict = "mu",
  centrality = "mean"
)
```

The posterior mean will not necessarily equal exactly 50 ms because of sampling variability. However, the 95% credible interval should contain the population value of 50 ms when the model successfully recovers the simulated effect.

#### Estimating the Exponential Component

Condition-specific estimates of $\tau$ can be obtained by requesting predictions for the `tau` distributional parameter.

```{r}
#| label: estimate-tau-classical
#| echo: false
# Classical model: per-condition tau (the exponential component).
tau_estimates <- estimate_means(
  fit_brms_new,
  by = "Blur",
  predict = "tau",
  centrality = "mean"
)
```

@tbl-classical-means shows the Gaussian-location and exponential-component estimates together, by condition.

```{r}
#| label: build-tbl-classical-means
#| echo: false
classical_means_mu <- as.data.frame(mu_estimates)
classical_means_tau <- as.data.frame(tau_estimates)

classical_means_tbl <- rbind(
  data.frame(
    Blur = classical_means_mu$Blur,
    Parameter = "μ",
    Mean = fmt_num(classical_means_mu$Mu),
    `95% CrI` = fmt_ci(classical_means_mu$CI_low, classical_means_mu$CI_high),
    pd = fmt_pd(classical_means_mu$pd),
    check.names = FALSE
  ),
  # estimate_means(predict = "tau") names the estimate column "Mean", not
  # "Tau" (unlike "mu"/"beta", which do get their own name), with this
  # model's default marginaleffects backend.
  data.frame(
    Blur = classical_means_tau$Blur,
    Parameter = "τ",
    Mean = fmt_num(classical_means_tau$Mean),
    `95% CrI` = fmt_ci(classical_means_tau$CI_low, classical_means_tau$CI_high),
    pd = fmt_pd(classical_means_tau$pd),
    check.names = FALSE
  )
)
classical_means_tbl <- classical_means_tbl[order(classical_means_tbl$Blur), ]
rownames(classical_means_tbl) <- NULL
classical_means_tbl$Blur <- as.character(classical_means_tbl$Blur)
classical_means_tbl$Blur[duplicated(classical_means_tbl$Blur)] <- ""
```

```{r}
#| label: tbl-classical-means
#| tbl-cap: "Estimated Gaussian-location ($\\mu$) and exponential ($\\tau$) parameters, by condition, from the classical model."
#| echo: false
#| results: asis
tt(classical_means_tbl)
```

The estimates should be close to 100 ms for Clear and 50 ms for Low-Blur for $\tau$.

The corresponding contrast can be obtained using `estimate_contrasts()`.

```{r}
#| label: contrast-tau-classical
#| echo: false
# Classical model: Blur contrast in tau (the exponential component).
tau_difference <- estimate_contrasts(
  fit_brms_new,
  contrast = "Blur",
  predict = "tau",
  centrality = "mean"
)
```

```{r}
#| label: build-tbl-classical-contrast
#| echo: false
classical_contrast_mu <- as.data.frame(mu_difference)
classical_contrast_tau <- as.data.frame(tau_difference)

classical_contrast_tbl <- rbind(
  data.frame(
    Contrast = paste(
      classical_contrast_mu$Level1,
      "-",
      classical_contrast_mu$Level2
    ),
    Parameter = "μ",
    Mean = fmt_num(classical_contrast_mu$Mean),
    `95% CrI` = fmt_ci(
      classical_contrast_mu$CI_low,
      classical_contrast_mu$CI_high
    ),
    pd = fmt_pd(classical_contrast_mu$pd),
    check.names = FALSE
  ),
  data.frame(
    Contrast = paste(
      classical_contrast_tau$Level1,
      "-",
      classical_contrast_tau$Level2
    ),
    Parameter = "τ",
    Mean = fmt_num(classical_contrast_tau$Mean),
    `95% CrI` = fmt_ci(
      classical_contrast_tau$CI_low,
      classical_contrast_tau$CI_high
    ),
    pd = fmt_pd(classical_contrast_tau$pd),
    check.names = FALSE
  )
)
```

```{r}
#| label: tbl-classical-contrast
#| tbl-cap: "Blur contrast in the Gaussian-location ($\\mu$) and exponential ($\\tau$) parameters, from the classical model."
#| echo: false
#| results: asis
tt(classical_contrast_tbl)
```

The results seen in @tbl-classical-contrast are very similar to the default paramaterization for $\tau$, but unlike the default model, the classical model's $\mu$ contrast directly recovers the simulated 50-ms shift in the Gaussian location.

This example illustrates the central difference between the two parameterizations. The default {**`brms`**} ex-Gaussian model directly estimates the expected response and labels that parameter `mu`. The exponential component is labeled `beta`, although it corresponds to the classical $\tau$ parameter. In contrast, the `{cogmod}`'s `rt_exgaussian()` model uses `mu` for the Gaussian-location parameter and `tau` for the exponential mean, consistent with the classical parameterization commonly used in reaction-time research.

@fig-abstract summarizes this simulation in a single picture: the two simulated distributions differ by 50 ms in the Gaussian location $\mu$, but only the classical parameterization's contrast on $\mu$ detects that difference — the default {brms} model's contrast (on $E(RT)$) does not.

```{r}
#| label: fig-abstract
#| fig-cap: "Graphical summary of the Simulation Example. Top: simulated reaction-time distributions for the Clear and Low-Blur conditions, which differ by 50 ms in the Gaussian location $\\mu$. Bottom: the Blur contrast under each parameterization, with credibility determined by whether the 95% credible interval excludes zero; pd is the probability of direction."
#| echo: false
#| fig-width: 7
#| fig-height: 6.5

mu_lookup <- data.frame(Blur = c("Clear", "Low-Blur"), mu = sim_params$mu)

# Single (non-faceted) panel with both densities overlapping, so the Delta-mu
# arrow can span between them without ggplot's per-facet annotate() duplication.
p_dist <- ggplot(d, aes(x = rt, fill = Blur, colour = Blur)) +
  ggdist::stat_slab(alpha = 0.55, linewidth = 0, normalize = "panels") +
  geom_segment(
    data = mu_lookup, aes(x = mu, xend = mu, y = 0, yend = 1.05, colour = Blur),
    linetype = "dashed", linewidth = 0.7, show.legend = FALSE
  ) +
  annotate(
    "segment",
    x = min(mu_lookup$mu), xend = max(mu_lookup$mu), y = 1.15, yend = 1.15,
    arrow = grid::arrow(length = grid::unit(1.6, "mm"), ends = "both"),
    linewidth = 0.6, colour = "grey20"
  ) +
  annotate(
    "text",
    x = mean(mu_lookup$mu), y = 1.25,
    label = paste0("Δμ = ", diff(range(mu_lookup$mu)), " ms"),
    size = 4.4, fontface = "bold", colour = "grey20"
  ) +
  scale_fill_manual(values = c("Clear" = "darkgreen", "Low-Blur" = "darkred")) +
  scale_colour_manual(values = c("Clear" = "darkgreen", "Low-Blur" = "darkred")) +
  coord_cartesian(xlim = c(300, 900), ylim = c(0, 1.4), clip = "off") +
  labs(x = "Reaction time (ms)", y = NULL, title = "Simulated Distributions (μ differs by 50 ms)", fill = NULL, colour = NULL) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid = element_blank(),
    legend.position = "top",
    plot.title = element_text(face = "bold", hjust = 0.5, size = 12)
  )

# Verdict boxes: pull the actual Blur contrast from each fitted model rather
# than hardcoding the estimate/CI/credibility.
default_mu_contrast   <- as.data.frame(overall_difference)
classical_mu_contrast <- as.data.frame(mu_difference)

# pd >= .975 is the conventional threshold (roughly equivalent to a two-sided
# 95% credible interval excluding zero) for treating a direction as credible.
pd_threshold <- 0.975
is_credible <- function(pd) pd >= pd_threshold
verdict_label <- function(est, pd) {
  if (pd < pd_threshold) "UNCERTAIN" else if (est > 0) "EFFECT ABOVE 0" else "EFFECT BELOW 0"
}

make_verdict_box <- function(title, est, ci_low, ci_high, pd, credible) {
  txt <- paste0(
    title, "\n",
    "Δ = ", fmt_num(est), " ms   95% CrI ", fmt_ci(ci_low, ci_high), "   pd = ", fmt_pd(pd), "\n",
    verdict_label(est, pd)
  )
  fill_colour <- if (credible) "#d4edda" else "#f0f0f0"
  ggplot() +
    annotate("rect", xmin = 0, xmax = 1, ymin = 0, ymax = 1, fill = fill_colour, colour = "grey30", linewidth = 0.8) +
    annotate("text", x = 0.5, y = 0.5, label = txt, size = 3.8, fontface = "bold", lineheight = 1.3, colour = "grey10") +
    theme_void() +
    coord_cartesian(xlim = c(0, 1), ylim = c(0, 1), expand = FALSE, clip = "off")
}

p_default_box <- make_verdict_box(
  "Default {brms}\n(μ = E(RT))",
  default_mu_contrast$Mean, default_mu_contrast$CI_low, default_mu_contrast$CI_high, default_mu_contrast$pd,
  is_credible(default_mu_contrast$pd)
)

p_classical_box <- make_verdict_box(
  "Classical ex-Gaussian\n(μ = Gaussian location)",
  classical_mu_contrast$Mean, classical_mu_contrast$CI_low, classical_mu_contrast$CI_high, classical_mu_contrast$pd,
  is_credible(classical_mu_contrast$pd)
)

p_banner <- ggplot() +
  annotate("rect", xmin = 0, xmax = 1, ymin = 0, ymax = 1, fill = "#fff3cd", colour = "grey30", linewidth = 0.8) +
  annotate(
    "text", x = 0.5, y = 0.5,
    label = "The default {brms} family does not capture the difference in μ!",
    size = 5, fontface = "bold", colour = "#664d03"
  ) +
  theme_void() +
  coord_cartesian(xlim = c(0, 1), ylim = c(0, 1), expand = FALSE)

p_dist / (p_default_box | p_classical_box) / p_banner +
  patchwork::plot_layout(heights = c(2.2, 0.8, 0.5))
```

## Real Example

In order to demonstrate how the difference between parameterizations can affect the interpretation of real data, we re-analyze Experiment 1 from @angele2023. The main focus of @angele2023's study was to test whether the masked priming paradigm can be successfully implemented in a Javascript-based online experiment. However, @angele2023 also attempt to make a theoretical contribution by using the ex-Gaussian model to evaluate @forster1998pros's savings hypothesis. @forster1998pros argued that, if the prime is identical to the target, having pre-processed the prime will allow participants to respond faster. In this case, the whole distribution of reaction times should be shifted to the left for the related condition compared to the unrelated condition, i.e., there should be an effect on $\mu$, rather than on the slower responses, i.e., $\tau$. The experiment had a more complex design than the previous examples: participants completed a masked identity priming lexical decision task across two prime exposure durations (33 ms vs. 50 ms) and two priming conditions (Identical vs. Unrelated).

To fit their ex-Gaussian models, @angele2023 used the built-in `exgaussian()` family in {brms}. As established above, the primary formula in default {brms} models the overall distributional mean $E(RT) = \mu_{\text{Gaussian}} + \tau$, despite assigning the label `mu` to this parameter. In their manuscript, @angele2023 interpreted regression effects on `mu` as changes in the location of the Gaussian component $\mu$.

Here, we apply the classical ex-Gaussian parameterization using `{cogmod}`, modeling reaction times with maximal random intercepts and random slopes for participants (`source`) and items (`Target`) across both $\mu$ (Gaussian location) and $\tau$ (exponential mean). Reaction time is coded in milliseconds in this dataset. As discussed above, we fit this model with `link_mu = "identity"` and `link_sigma = link_tau = "log"` rather than {cogmod}'s default softplus links, so priors and raw model coefficients for $\mu$ are directly on the millisecond scale, and priors for $\sigma$ and $\tau$ are on the (familiar, scale-robust) log scale. As with the simulated example, all condition-specific estimates and contrasts are still obtained through `estimate_means()`/`estimate_contrasts()` rather than by reading the raw coefficients directly.

Prior specification for the classical model is summarized in @tbl-priors-real.

```{r}
#| label: priors-real-exp1
#| eval: false
#| echo: false
priors_classical <- c(
  # Gaussian location mu: Intercept & effects (identity link, ms scale)
  prior(normal(500, 100), class = "Intercept"),
  prior(normal(0, 100), class = "b", coef = "Condition1"),
  prior(normal(0, 100), class = "b", coef = "PrimeDuration1"),
  prior(normal(0, 100), class = "b", coef = "Condition1:PrimeDuration1"),

  # Exponential tail tau (on log scale): Intercept & effects
  prior(normal(log(100), 0.5), class = "Intercept", dpar = "tau"),
  prior(normal(0, 0.5), class = "b", dpar = "tau"),

  # Gaussian SD sigma (on log scale): Intercept
  prior(normal(log(60), 0.5), class = "Intercept", dpar = "sigma")
)
```

```{r}
#| label: tbl-priors-real
#| tbl-cap: "Prior specification for the classical ex-Gaussian model fit to Experiment 1 from @angele2023."
#| echo: false
#| results: asis
priors_table_real <- data.frame(
  Parameter = c(
    "Intercept (μ)",
    "Condition (μ)",
    "Prime Duration (μ)",
    "Condition x Prime Duration (μ)",
    "Intercept (τ)",
    "All effects (τ)",
    "Intercept (σ)"
  ),
  Prior = c(
    "Normal(500, 100)",
    "Normal(0, 100)",
    "Normal(0, 100)",
    "Normal(0, 100)",
    "Normal(log(100), 0.5)",
    "Normal(0, 0.5)",
    "Normal(log(60), 0.5)"
  )
)
tt(priors_table_real)
```

### Initial Values Specification

When fitting complex hierarchical ex-Gaussian models with random effects on parameters constrained to be positive (such as $\tau$ and $\sigma$), Stan's default unconstrained random initialization (in $[-2, 2]$) can cause severe numerical instability. Standard deviations randomly drawn near the upper end of this range can generate extreme individual-level parameter deviations, causing MCMC chains to immediately reject initial values (`log probability evaluates to log(0)`) or fail to start.

To prevent initial value rejection and ensure reliable sampling across all chains, we supply an explicit initialization function (`init_classical`) that sets reasonable starting values for all model parameters. The full code for this function is omitted here for brevity but can be found in the manuscript's `.qmd` source file on GitHub.

```{r}
#| label: inits-real-exp1
#| eval: false
#| echo: false

# Subset data to determine random effect dimensions
data_fit <- exp1_data_to_include |>
  filter(corr == 1 & StimulusType == "Word" & rt > 250 & rt < 1800)

N_source <- length(unique(data_fit$source))
N_target <- length(unique(data_fit$Target))

# Function generating stable initial values for all Stan parameters
init_classical <- function() {
  list(
    Intercept = 500, # Gaussian location (ms, identity link)
    Intercept_sigma = log(60), # Gaussian SD (log link)
    Intercept_tau = log(100), # Exponential tail (log link)
    b = rep(0, 3), # Fixed effects on mu
    b_tau = rep(0, 3), # Fixed effects on tau
    sd_1 = rep(30, 4), # Participant SDs for mu (ms)
    sd_2 = rep(30, 4), # Item SDs for mu (ms)
    sd_3 = rep(0.1, 4), # Participant SDs for tau (small to prevent overflow)
    sd_4 = rep(0.1, 4), # Item SDs for tau (small to prevent overflow)
    z_1 = matrix(0, nrow = 4, ncol = N_source), # Participant random effects for mu
    z_2 = matrix(0, nrow = 4, ncol = N_target), # Item random effects for mu
    z_3 = matrix(0, nrow = 4, ncol = N_source), # Participant random effects for tau
    z_4 = matrix(0, nrow = 4, ncol = N_target), # Item random effects for tau
    L_1 = diag(4), # Cholesky factor for participant random effects (mu)
    L_2 = diag(4), # Cholesky factor for item random effects (mu)
    L_3 = diag(4), # Cholesky factor for participant random effects (tau)
    L_4 = diag(4) # Cholesky factor for item random effects (tau)
  )
}
```

The classical ex-Gaussian model for Experiment 1 is specified and fit as follows:

```{r}
#| label: fit-real-exp1-code
#| eval: false
# Fit classical ex-Gaussian hierarchical model to Experiment 1 RT data
blmm_exp1_classical_rt <- brm(
  data = data_fit,
  formula = bf(
    rt ~ Condition *
      PrimeDuration +
      (1 + Condition * PrimeDuration | source) +
      (1 + Condition * PrimeDuration | Target),
    tau ~ Condition *
      PrimeDuration +
      (1 + Condition * PrimeDuration | source) +
      (1 + Condition * PrimeDuration | Target),
    sigma ~ 1
  ),
  warmup = 1000,
  iter = 5000,
  chains = 4,
  prior = priors_classical,
  family = rt_exgaussian(link = "identity", link_tau = "log", link_sigma = "log"), # {cogmod} family with identity/log links
  stanvars = rt_exgaussian_stanvars(), # Stan code injected into the model
  inits = init_classical,
  cores = 4,
  backend = "cmdstanr",
  threads = threading(2)
)
```

Depending on the hardware used, this model may take more than 30 minutes to fit. Because of this, it is recommended to fit it once and then save the fitted model object to a file for later use.

### Estimating Ex-Gaussian Parameters with `{easystats}`

As in the Simulation Example, we use `{modelbased}`'s `estimate_means()` and `estimate_contrasts()` to obtain condition-specific posterior estimates and contrasts, rather than computing them by hand from the raw posterior draws. Because this model has thousands of posterior draws and a large number of by-participant and by-item random effects, we set `backend = "emmeans"` (instead of the default `marginaleffects` backend) to keep the marginal computations memory-efficient.

```{r}
#| label: fit-real-exp1
#| echo: false
#| message: false
#| warning: false

exp1_qs2_path <- if (file.exists("Angele_et_al_2022/blmm_exp1_classical_rt.qs2")) {
  "Angele_et_al_2022/blmm_exp1_classical_rt.qs2"
} else if (file.exists("../Angele_et_al_2022/blmm_exp1_classical_rt.qs2")) {
  "../Angele_et_al_2022/blmm_exp1_classical_rt.qs2"
} else {
  here::here("Angele_et_al_2022", "blmm_exp1_classical_rt.qs2")
}

fit_exp1_classical <- qs_read(exp1_qs2_path)

# Harmonizes contrast direction to "Unrelated - Identical" / "50 ms - 33 ms",
# matching the sign of the raw population-level regression coefficients.
flip_contrast <- function(x) {
  d <- as.data.frame(x)
  new_low <- -d$CI_high
  new_high <- -d$CI_low
  d$Difference <- -d$Difference
  d$CI_low <- new_low
  d$CI_high <- new_high
  d
}
```

```{r}
#| label: estimate-means-real
#| echo: false

real_means_mu <- as.data.frame(estimate_means(
  fit_exp1_classical,
  by = c("Condition", "PrimeDuration"),
  predict = "mu",
  centrality = "mean",
  backend = "emmeans"
))
real_means_tau <- as.data.frame(estimate_means(
  fit_exp1_classical,
  by = c("Condition", "PrimeDuration"),
  predict = "tau",
  centrality = "mean",
  backend = "emmeans"
))

real_means_tbl <- rbind(
  data.frame(
    Condition = real_means_mu$Condition,
    PrimeDuration = real_means_mu$PrimeDuration,
    Parameter = "μ",
    Mean = fmt_num(real_means_mu$Mu),
    `95% CrI` = fmt_ci(real_means_mu$CI_low, real_means_mu$CI_high),
    pd = fmt_pd(real_means_mu$pd),
    check.names = FALSE
  ),
  data.frame(
    Condition = real_means_tau$Condition,
    PrimeDuration = real_means_tau$PrimeDuration,
    Parameter = "τ",
    Mean = fmt_num(real_means_tau$Tau),
    `95% CrI` = fmt_ci(real_means_tau$CI_low, real_means_tau$CI_high),
    pd = fmt_pd(real_means_tau$pd),
    check.names = FALSE
  )
)
real_means_tbl <- real_means_tbl[
  order(real_means_tbl$Condition, real_means_tbl$PrimeDuration),
]
rownames(real_means_tbl) <- NULL
real_means_tbl$Condition <- as.character(real_means_tbl$Condition)
real_means_tbl$PrimeDuration <- as.character(real_means_tbl$PrimeDuration)
dup_duration <- duplicated(real_means_tbl[c("Condition", "PrimeDuration")])
dup_condition <- duplicated(real_means_tbl$Condition)
real_means_tbl$PrimeDuration[dup_duration] <- ""
real_means_tbl$Condition[dup_condition] <- ""
```

```{r}
#| label: tbl-real-means
#| tbl-cap: "Estimated Gaussian-location ($\\mu$) and exponential ($\\tau$) parameters by condition and prime duration for Experiment 1 from @angele2023 under the classical model."
#| echo: false
#| results: asis
tt(real_means_tbl)
```

@tbl-real-means displays the estimated posterior means for $\mu$ and $\tau$ across priming conditions and exposure durations. @tbl-real-contrasts below presents the model's population-level effects (main effect of relatedness, main effect of prime duration, and their interaction) alongside simple priming contrasts (Unrelated minus Identical) at each prime duration, for both parameters side by side.

```{r}
#| label: estimate-contrasts-real
#| echo: false
#| cache: true

mu_main_cond <- flip_contrast(estimate_contrasts(
  fit_exp1_classical,
  contrast = "Condition",
  predict = "mu",
  centrality = "mean",
  backend = "emmeans"
))
mu_main_dur <- flip_contrast(estimate_contrasts(
  fit_exp1_classical,
  contrast = "PrimeDuration",
  predict = "mu",
  centrality = "mean",
  backend = "emmeans"
))
mu_simple <- flip_contrast(estimate_contrasts(
  fit_exp1_classical,
  contrast = "Condition",
  by = "PrimeDuration",
  predict = "mu",
  centrality = "mean",
  backend = "emmeans"
))

tau_main_cond <- flip_contrast(estimate_contrasts(
  fit_exp1_classical,
  contrast = "Condition",
  predict = "tau",
  centrality = "mean",
  backend = "emmeans"
))
tau_main_dur <- flip_contrast(estimate_contrasts(
  fit_exp1_classical,
  contrast = "PrimeDuration",
  predict = "tau",
  centrality = "mean",
  backend = "emmeans"
))
tau_simple <- flip_contrast(estimate_contrasts(
  fit_exp1_classical,
  contrast = "Condition",
  by = "PrimeDuration",
  predict = "tau",
  centrality = "mean",
  backend = "emmeans"
))

# mu is estimated on the identity scale, so the raw population-level
# interaction coefficient already is the response-scale (ms) interaction.
fixed_effects_real <- as.data.frame(parameters::model_parameters(
  fit_exp1_classical,
  effects = "fixed",
  centrality = "mean"
))
mu_intx <- fixed_effects_real[
  fixed_effects_real$Parameter == "b_Condition1:PrimeDuration1",
]

# tau uses a nonlinear link, so its response-scale (ms) interaction is not
# the raw (log-scale) regression coefficient. estimate_contrasts() has no
# built-in "difference of differences" comparison, so we get the four cell-wise
# posterior draws from estimate_means() (keep_iterations = TRUE) and combine
# them ourselves -- the draws themselves still come from {modelbased}/{easystats}.
tau_means_draws <- as.data.frame(estimate_means(
  fit_exp1_classical,
  by = c("Condition", "PrimeDuration"),
  predict = "tau",
  centrality = "mean",
  backend = "emmeans",
  keep_iterations = TRUE
))
iter_cols <- grep("^iter_", names(tau_means_draws), value = TRUE)
get_draws <- function(condition, duration) {
  as.numeric(tau_means_draws[
    tau_means_draws$Condition == condition &
      tau_means_draws$PrimeDuration == duration,
    iter_cols
  ])
}
tau_intx_draws <- (get_draws("Unrelated", "50 ms") -
  get_draws("Identical", "50 ms")) -
  (get_draws("Unrelated", "33 ms") - get_draws("Identical", "33 ms"))
tau_intx <- data.frame(
  Mean = mean(tau_intx_draws),
  CI_low = quantile(tau_intx_draws, 0.025)[[1]],
  CI_high = quantile(tau_intx_draws, 0.975)[[1]],
  # tau_intx isn't from estimate_contrasts()/model_parameters(), so pd is
  # computed directly from its draws via {bayestestR} (part of {easystats}).
  pd = as.numeric(bayestestR::p_direction(tau_intx_draws))
)

# Effects as originally reported by @angele2023 under their default-brms
# E(RT) parameterization (their Table 3); we did not refit their default model.
default_ert <- c(
  "31.5 [27.7, 35.3]",
  "-10.9 [-14.5, -7.3]",
  "19.3 [12.6, 26.0]",
  "~22.0",
  "~41.0"
)

real_contrasts_tbl <- data.frame(
  Effect = c(
    "Main Effect of Relatedness",
    "Main Effect of Prime Duration",
    "Relatedness x Prime Duration Interaction",
    paste("Priming Effect at", mu_simple$PrimeDuration)
  ),
  `Default brms E(RT)` = default_ert,
  Mu = fmt_num(c(
    mu_main_cond$Difference,
    mu_main_dur$Difference,
    mu_intx$Mean,
    mu_simple$Difference
  )),
  `Mu 95% CrI` = fmt_ci(
    c(
      mu_main_cond$CI_low,
      mu_main_dur$CI_low,
      mu_intx$CI_low,
      mu_simple$CI_low
    ),
    c(
      mu_main_cond$CI_high,
      mu_main_dur$CI_high,
      mu_intx$CI_high,
      mu_simple$CI_high
    )
  ),
  `Mu pd` = fmt_pd(c(
    mu_main_cond$pd,
    mu_main_dur$pd,
    mu_intx$pd,
    mu_simple$pd
  )),
  Tau = fmt_num(c(
    tau_main_cond$Difference,
    tau_main_dur$Difference,
    tau_intx$Mean,
    tau_simple$Difference
  )),
  `Tau 95% CrI` = fmt_ci(
    c(
      tau_main_cond$CI_low,
      tau_main_dur$CI_low,
      tau_intx$CI_low,
      tau_simple$CI_low
    ),
    c(
      tau_main_cond$CI_high,
      tau_main_dur$CI_high,
      tau_intx$CI_high,
      tau_simple$CI_high
    )
  ),
  `Tau pd` = fmt_pd(c(
    tau_main_cond$pd,
    tau_main_dur$pd,
    tau_intx$pd,
    tau_simple$pd
  )),
  check.names = FALSE
)

# Named lookup so in-text numbers are always pulled from the table above
# rather than typed by hand.
get_val <- function(effect, col) {
  real_contrasts_tbl[real_contrasts_tbl$Effect == effect, col]
}

# Display-only: long format (one row per Effect x Parameter) instead of Mu/Tau
# side by side. get_val() above still reads the wide table, so in-text prose
# is unaffected.
real_contrasts_display <- rbind(
  data.frame(
    Effect = real_contrasts_tbl$Effect,
    `Default brms E(RT)` = real_contrasts_tbl$`Default brms E(RT)`,
    `Classical Parameter` = "μ",
    `Mean [95% CrI]` = paste(
      real_contrasts_tbl$Mu,
      real_contrasts_tbl$`Mu 95% CrI`
    ),
    pd = real_contrasts_tbl$`Mu pd`,
    check.names = FALSE
  ),
  data.frame(
    Effect = real_contrasts_tbl$Effect,
    `Default brms E(RT)` = real_contrasts_tbl$`Default brms E(RT)`,
    `Classical Parameter` = "τ",
    `Mean [95% CrI]` = paste(
      real_contrasts_tbl$Tau,
      real_contrasts_tbl$`Tau 95% CrI`
    ),
    pd = real_contrasts_tbl$`Tau pd`,
    check.names = FALSE
  )
)
real_contrasts_display$Effect <- factor(
  real_contrasts_display$Effect,
  levels = real_contrasts_tbl$Effect
)
real_contrasts_display <- real_contrasts_display[
  order(real_contrasts_display$Effect),
]
real_contrasts_display$Effect <- as.character(real_contrasts_display$Effect)
# Default brms E(RT) is an Effect-level (not Parameter-level) value, so it's
# identical across the mu/tau row pair for each effect; blank the repeat.
dup_effect <- duplicated(real_contrasts_display$Effect)
real_contrasts_display$`Default brms E(RT)`[dup_effect] <- ""
rownames(real_contrasts_display) <- NULL
```

```{r}
#| label: tbl-real-contrasts
#| tbl-cap: "Fixed effects and priming contrasts in the Gaussian-location ($\\mu$) and exponential ($\\tau$) parameters for Experiment 1 from @angele2023 under the classical model, alongside the default-brms E(RT) effects as originally reported by @angele2023. All estimates are in milliseconds and reflect Unrelated minus Identical (or 50 ms minus 33 ms) differences; pd is the probability of direction."
#| echo: false
#| results: asis
tt(real_contrasts_display)
```

As shown in @tbl-real-contrasts, under the classical parameterization, the true Gaussian location $\mu$ shows a robust overall main effect of relatedness (`r get_val("Main Effect of Relatedness", "Mu")` ms, 95% CrI `r get_val("Main Effect of Relatedness", "Mu 95% CrI")`). Unlike the effect reported as $\mu$ under the default parameterization, the main effect of prime duration on $\mu$ is non-credible under the classical parameterization (`r get_val("Main Effect of Prime Duration", "Mu")` ms, 95% CrI `r get_val("Main Effect of Prime Duration", "Mu 95% CrI")`, with the 95% CrI narrowly containing zero). The model also reveals a credible interaction between relatedness and prime duration (`r get_val("Relatedness x Prime Duration Interaction", "Mu")` ms, 95% CrI `r get_val("Relatedness x Prime Duration Interaction", "Mu 95% CrI")`), indicating that the priming effect on $\mu$ was larger at 50 ms (`r get_val("Priming Effect at 50 ms", "Mu")` ms, 95% CrI `r get_val("Priming Effect at 50 ms", "Mu 95% CrI")`) than at 33 ms (`r get_val("Priming Effect at 33 ms", "Mu")` ms, 95% CrI `r get_val("Priming Effect at 33 ms", "Mu 95% CrI")`).

For the exponential component $\tau$, the main effect of relatedness is small and non-credible (`r get_val("Main Effect of Relatedness", "Tau")` ms, 95% CrI `r get_val("Main Effect of Relatedness", "Tau 95% CrI")`), as is the interaction term (`r get_val("Relatedness x Prime Duration Interaction", "Tau")` ms, 95% CrI `r get_val("Relatedness x Prime Duration Interaction", "Tau 95% CrI")`). Consequently, the priming effect (i.e., the difference between related and unrelated prime conditions) on $\tau$ is non-credible at 33 ms (`r get_val("Priming Effect at 33 ms", "Tau")` ms, 95% CrI `r get_val("Priming Effect at 33 ms", "Tau 95% CrI")`) and 50 ms (`r get_val("Priming Effect at 50 ms", "Tau")` ms, 95% CrI `r get_val("Priming Effect at 50 ms", "Tau 95% CrI")`), consistent with Forster's [-@forster1998pros] savings hypothesis that would expect such an effect to show in $\mu$ rather than in $\tau$. However, the model reveals a credible main effect of prime duration on $\tau$ (`r get_val("Main Effect of Prime Duration", "Tau")` ms, 95% CrI `r get_val("Main Effect of Prime Duration", "Tau 95% CrI")`), showing that increasing prime duration from 33 ms to 50 ms moderately shortened the right tail.

### Comparison with Angele et al. (2023)

Comparing these classical ex-Gaussian estimates with those originally reported by @angele2023 (the "Default brms E(RT)" column of @tbl-real-contrasts) highlights the practical consequences of model parameterization for theoretical inferences.

1.  *Parameter Interpretation*. @angele2023 reported fixed effects for relatedness, prime duration, and their interaction from default {brms} under the assumption that `mu` represented the Gaussian location component $\mu_{\text{Gaussian}}$. Actually, the `brms` default `mu` estimates $E(RT) = \mu_{\text{Gaussian}} + \tau$. Consequently, @angele2023 interpreted overall differences in the distribution mean $E(RT)$ as changes in the Gaussian location $\mu_{\text{Gaussian}}$, when in fact they were a composite of changes in both $\mu_{\text{Gaussian}}$ and $\tau$.

2.  **Prime Duration Effects**: While @angele2023's interpretation of most of the effects does not differ substantially from what the classical parameterization results indicate, there is one interesting and potentially theoretically relevant difference: Based on their default {brms} model, @angele2023 concluded that increasing prime exposure duration from 33 ms to 50 ms produced a credible downward shift in the Gaussian location parameter $\mu$ ($b = -10.9$ ms, 95% CrI \[-14.5, -7.3\]). However, the classical parameterization reveals that the true main effect of prime duration on $\mu_{\text{Gaussian}}$ is not credible (`r get_val("Main Effect of Prime Duration", "Mu")` ms, 95% CrI `r get_val("Main Effect of Prime Duration", "Mu 95% CrI")`, with the 95% CrI containing zero). Instead, our reanalysis shows that the overall reduction in RTs at 50 ms prime duration was driven almost entirely by a compression of the exponential tail $\tau$ (`r get_val("Main Effect of Prime Duration", "Tau")` ms, 95% CrI `r get_val("Main Effect of Prime Duration", "Tau 95% CrI")`). The effect on $E(RT)$ reported by @angele2023 as a location shift in $\mu$ was actually the composite sum of a non-credible $\mu$ shift (`r get_val("Main Effect of Prime Duration", "Mu")` ms) and a credible $\tau$ tail reduction (`r get_val("Main Effect of Prime Duration", "Tau")` ms).

3.  **Priming Effects and Interaction**: For the priming manipulation, the main effect of relatedness on $\mu_{\text{Gaussian}}$ (`r get_val("Main Effect of Relatedness", "Mu")` ms, 95% CrI `r get_val("Main Effect of Relatedness", "Mu 95% CrI")`) and the interaction (`r get_val("Relatedness x Prime Duration Interaction", "Mu")` ms, 95% CrI `r get_val("Relatedness x Prime Duration Interaction", "Mu 95% CrI")`) closely mirror the $E(RT)$ effects reported by @angele2023 (main effect $\sim 31.5$ ms, interaction $\sim 19.3$ ms). The estimates for the two parameterizations align closely for the priming effect because $\tau$ showed no credible priming differences (`r get_val("Main Effect of Relatedness", "Tau")` ms, 95% CrI `r get_val("Main Effect of Relatedness", "Tau 95% CrI")`). When $\Delta \tau \approx 0$, changes in the distributional mean $\Delta E(RT)$ track changes in the Gaussian location $\Delta \mu$ nearly 1-to-1.

4.  **Summary**: This empirical comparison demonstrates that misinterpreting default {brms} parameters can lead to subtle misattributions regarding *which* distributional component is altered by an experimental manipulation. In @angele2023, a tail-shortening effect of prime duration ($\Delta \tau =$ `r get_val("Main Effect of Prime Duration", "Tau")` ms) was attributed to the Gaussian location parameter (`mu`). Adopting the classical parameterization showcases some uncertainty regarding the existence of an effect of prime duration on $\mu$ (pd = `r get_val("Main Effect of Prime Duration", "Mu pd")`), underscoring that parameterization affects not only which component an effect is attributed to, but the confidence with which its existence can be claimed at all.

## Conclusion

With this commentary, we hope to encourage researchers to use the appropriate ex-Gaussian parameterization when modeling reaction-time data with the {brms} package. The ex-Gaussian distribution provides a powerful framework for moving beyond analyses of the mean by separating differences in the central portion of the reaction-time distribution from differences in its right tail.

However, this flexibility also requires careful attention to model parameterization. In particular, researchers should not assume that parameters with familiar labels necessarily have the same interpretation across software implementations. Researchers should therefore examine the statistical definition and underlying implementation of a model before fitting it and interpreting its parameters. This is especially important when using flexible modeling software that may adopt parameterizations that differ from those commonly used in a particular research literature. We hope this commentary serves both as a practical guide to fitting the classical ex-Gaussian model in {brms} and as a broader reminder that convenient software defaults should not replace careful consideration of what a model actually estimates.

# References
